---
title: "Data Exploration"
subtitle: "Data Exploration"
author: "Martin Brand"
date: "2022-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE, results="hide", echo=TRUE}
library(ggplot2)
library(readxl)
library(patchwork)
library(tidyverse)
library(caret)
library(reshape2)
```

## Preprocessing of the Data
We import the data provided by the authors as "Supplementary Table 1" and load it into a dataframe. The dataset contains for example the MSI/MSS classifications. Additionally, we import the "DepMap_18Q4_data.rds" dataset, which contains all dependency scores. 
```{r, warning=FALSE}
# suppress warning message caused by na values
data_dir <- "C:/Users/Martin/Documents/Master/Forschungspraktikum/Chan/data/"
# read file with MSS/MSI classifications
data_suppl <- read_excel(paste0(data_dir, "supplementary_table.xlsx"))
# transform "NA" strings to na values
data_suppl[data_suppl=="NA"] <- NA 
# transform numbers saved as strings to numbers
data_suppl <- transform(data_suppl, DRIVE_WRN_D2=as.numeric(DRIVE_WRN_D2), CRISPR_WRN_CERES = as.numeric(CRISPR_WRN_CERES), avg_WRN_dep=as.numeric(avg_WRN_dep))
# read file with dependency scores
data_18Q4 <- read_rds(file.path(data_dir, 'DepMap_18Q4_data.rds'))
dep_scores_18Q4_drive <- data_18Q4[["DRIVE"]]
dep_scores_18Q4_achilles <- data_18Q4[["CRISPR"]]
```

## Visualization of Classifications
We visualize the ratio of cell lines classified as MSS cell and cell lines classified as MSI cells for both Achilles and DRIVE datasets together and separately. We see that the number of MSS cell lines is much higher than the number of MSI cell lines.
```{r}
# count rows of MSI and MSS classified cell lines
counts_msi <- sum(data_suppl$CCLE_MSI=="MSI" & (!is.na(data_suppl$DRIVE_WRN_D2) | !is.na(data_suppl$CRISPR_WRN_CERES)), na.rm = TRUE)
counts_mss <- sum(data_suppl$CCLE_MSI=="MSS" & (!is.na(data_suppl$DRIVE_WRN_D2) | !is.na(data_suppl$CRISPR_WRN_CERES)), na.rm = TRUE)

# create dataframe with counts
counts.df <- data.frame(counts=c(counts_msi, counts_mss), classification=c("MSI", "MSS"))

# define limits on y axis
y_max <- 600

# create plot for both datasets together
bar_chart <- ggplot(data=counts.df, aes(x=classification, y=counts)) +
  geom_bar(stat = "identity", width = 0.5) +
  ggtitle("Both Datasets in total")
bar_chart <- bar_chart + labs(x="",y="number of cell lines") + ylim(0,y_max)

# count rows of MSI and MSS classified cell lines in Achilles dataset
counts_msi_achilles <- sum(data_suppl$CCLE_MSI=="MSI" & !is.na(data_suppl$CRISPR_WRN_CERES), na.rm = TRUE)
counts_mss_achilles <- sum(data_suppl$CCLE_MSI=="MSS" & !is.na(data_suppl$CRISPR_WRN_CERES), na.rm = TRUE)
counts_achilles.df <- data.frame(counts=c(counts_msi_achilles, counts_mss_achilles), classification=c("MSI", "MSS"))

# create plot for Achilles dataset
bar_chart_achilles <- ggplot(data=counts_achilles.df, aes(x=classification, y=counts)) +
  geom_bar(stat = "identity", width = 0.5) +
  ggtitle("Achilles")
bar_chart_achilles <- bar_chart_achilles + labs(x="",y="number of cell lines") + ylim(0,y_max)

# count rows of MSI and MSS classified cell lines in DRIVE dataset
counts_msi_drive <- sum(data_suppl$CCLE_MSI=="MSI" & !is.na(data_suppl$DRIVE_WRN_D2), na.rm = TRUE)
counts_mss_drive <- sum(data_suppl$CCLE_MSI=="MSS" & !is.na(data_suppl$DRIVE_WRN_D2), na.rm = TRUE)
counts_drive.df <- data.frame(counts=c(counts_msi_drive, counts_mss_drive), classification=c("MSI", "MSS"))

# create plot for DRIVE dataset
bar_chart_drive <- ggplot(data=counts_drive.df, aes(x=classification, y=counts)) +
  geom_bar(stat = "identity", width = 0.5) +
  ggtitle("DRIVE")
bar_chart_drive <- bar_chart_drive + labs(x="",y="number of cell lines") + ylim(0,y_max)

# show plots
bar_chart + bar_chart_achilles + bar_chart_drive
```

## Histograms
We visualize the the difference between the dependency scores for MSS and MSI cells as histograms.
```{r}
# create subset for both DRIVE and Achilles datasets
dep_scores_drive <- data_suppl %>% select(1,4,5)
dep_scores_drive <- dep_scores_drive[!is.na(dep_scores_drive$DRIVE_WRN_D2),]
dep_scores_drive <- dep_scores_drive[!is.na(dep_scores_drive$CCLE_MSI),]

dep_scores_achilles<- data_suppl %>% select(1,4,6)
dep_scores_achilles <- dep_scores_achilles[!is.na(dep_scores_achilles$CRISPR_WRN_CERES),]
dep_scores_achilles <- dep_scores_achilles[!is.na(dep_scores_achilles$CCLE_MSI),]

# create different dataframes for MSS and MSI cells
dep_scores_drive_mss <- data.frame(dep_scores_drive[dep_scores_drive$CCLE_MSI=="MSS",])  
dep_scores_drive_msi <- data.frame(dep_scores_drive[dep_scores_drive$CCLE_MSI=="MSI",])
dep_scores_achilles_mss <- data.frame(dep_scores_achilles[dep_scores_achilles$CCLE_MSI=="MSS",])
dep_scores_achilles_msi <-data.frame(dep_scores_achilles[dep_scores_achilles$CCLE_MSI=="MSI",])
# merge all datasets?

# change column names
colnames(dep_scores_drive_mss)[3] <- "dep_score"
colnames(dep_scores_drive_msi)[3] <- "dep_score"
colnames(dep_scores_achilles_mss)[3] <- "dep_score"
colnames(dep_scores_achilles_msi)[3] <- "dep_score"

# create template plot for mss and msi cells of DRIVE dataset
plot_drive_mss <- ggplot(dep_scores_drive_mss, aes(x=dep_score))
plot_drive_msi <- ggplot(dep_scores_drive_msi, aes(x=dep_score))

# add limits in x axis
limits <- c(-3,1)

# create histograms
histogram_plot_drive_mss <- plot_drive_mss +
  geom_histogram(binwidth = 0.1) +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("DRIVE and MSS")

histogram_plot_drive_msi <- plot_drive_msi +
  geom_histogram(binwidth = 0.1) +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score")+
  ggtitle("DRIVE and MSI")

# do the same with Achilles dataset
plot_achilles_mss <- ggplot(dep_scores_achilles_mss, aes(x=dep_score))
plot_achilles_msi <- ggplot(dep_scores_achilles_msi, aes(x=dep_score))

histogram_plot_achilles_mss <- plot_achilles_mss +
  geom_histogram(binwidth = 0.1) +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score")+
  ggtitle("Achilles and MSS")

histogram_plot_achilles_msi <- plot_achilles_msi +
  geom_histogram(binwidth = 0.1) +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score")+
  ggtitle("Achilles and MSI")


histogram_plot_drive_mss + histogram_plot_drive_msi + histogram_plot_achilles_mss + histogram_plot_achilles_msi

# the offset is different in DRIVE and Achilles... why?
```

## Density Plots
We visualize the the difference between the dependency scores for MSS and MSI cells as density plots.
```{r}
# create density plots
density_plot_drive_mss <- plot_drive_mss +
  geom_density() +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("DRIVE and MSS")

density_plot_drive_msi <- plot_drive_msi +
  geom_density() +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score")+
  ggtitle("DRIVE and MSI")

density_plot_achilles_mss <- plot_achilles_mss +
  geom_density() +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("Achilles and MSS")

density_plot_achilles_msi <- plot_achilles_msi +
  geom_density() +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("Achilles and MSI")

# show plots
density_plot_drive_mss + density_plot_drive_msi + density_plot_achilles_mss + density_plot_achilles_msi
```

## Box Plots
We visualize the the difference between the dependency scores for MSS and MSI cells as box plots.
```{r}
# create box plots
boxplot_drive_mss <- plot_drive_mss +
  geom_boxplot() +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("DRIVE and MSS")

boxplot_drive_msi <- plot_drive_msi +
  geom_boxplot() +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score")+
  ggtitle("DRIVE and MSI")

boxplot_achilles_mss <- plot_achilles_mss +
  geom_boxplot() +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("Achilles and MSS")

boxplot_achilles_msi <- plot_achilles_msi +
  geom_boxplot() +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("Achilles and MSI")

# show plots
boxplot_drive_mss + boxplot_drive_msi + boxplot_achilles_mss + boxplot_achilles_msi
```

## Correlation of Achilles and DRIVE dataset
```{r}
# extract dependency scores
dep_scores <- data_suppl %>% select(5,6)
# remove rows with na
dep_scores <- na.omit(dep_scores)
# # normalise values
# process <- preProcess(dep_scores, method=c("range"))
# norm_scores <- predict(process, dep_scores)

# remove outliers
outliers_achilles <- boxplot(dep_scores$CRISPR_WRN_CERES, plot=FALSE)$out
cleaned_dep_scores<- dep_scores[-which(dep_scores$CRISPR_WRN_CERES %in% outliers_achilles),]
outliers_drive <- boxplot(cleaned_dep_scores$DRIVE_WRN_D2, plot=FALSE)$out
cleaned_dep_scores<- cleaned_dep_scores[-which(cleaned_dep_scores$DRIVE_WRN_D2 %in% outliers_drive),]
# create scatter plot with and without outliers
corr_plot <- ggplot(dep_scores, aes(x=DRIVE_WRN_D2, y=CRISPR_WRN_CERES)) +
  geom_point()
cleaned_corr_plot <- ggplot(cleaned_dep_scores, aes(x=DRIVE_WRN_D2, y=CRISPR_WRN_CERES)) +
  geom_point()
corr_plot + cleaned_corr_plot
```

## Distribution of randomly picked KD/KO genes
For comparison we plot the distribution of randomly picked KO/KD genes.
```{r}
# pick randomly a gene  and extract dependency scores for this specific gene
abhd1_dep_scores_achilles <- data.frame(dep_scores_18Q4_achilles[,"ABHD1"])
# format dataset and add column for CCLE_ID
abhd1_dep_scores_achilles$CCLE_ID <- rownames(abhd1_dep_scores_achilles)
colnames(abhd1_dep_scores_achilles) <- c("dep_score_abhd1","CCLE_ID")
rownames(abhd1_dep_scores_achilles) <- c()
dep_scores_achilles_merged <- merge(abhd1_dep_scores_achilles, dep_scores_achilles, by="CCLE_ID")
dep_scores_achilles_merged <- dep_scores_achilles_merged[,c(1,3,2,4)]
colnames(dep_scores_achilles_merged) <- c("CCLE_ID","CCLE_MSI","ABHD1", "WRN")
dep_scores_achilles_merged <- melt(dep_scores_achilles_merged)
colnames(dep_scores_achilles_merged) <- c("CCLE_ID","CCLE_MSI","KD_KO_gene", "dep_score")

# create density plots for MSS cells
p1 <- ggplot(data=dep_scores_achilles_merged[dep_scores_achilles_merged$CCLE_MSI=="MSS",],aes(x=dep_score, color=KD_KO_gene)) +
  geom_density() +
  ggtitle("MSS and Achilles") +
  xlab("Dependency Score") +
  theme(legend.position="bottom")

# create density plots for MSI cells
p2 <- ggplot(data=dep_scores_achilles_merged[dep_scores_achilles_merged$CCLE_MSI=="MSI",],aes(x=dep_score, color=KD_KO_gene)) +
  geom_density() +
  ggtitle("MSI and Achilles") +
  xlab("Dependency Score") +
  theme(legend.position="bottom")

# plot results
p1 + p2
```