---
title: "Data Exploration"
subtitle: "Data Exploration"
author: "Martin Brand"
date: "2022-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE, results="hide", echo=TRUE}
library(ggplot2)
library(readxl)
library(patchwork)
```

We import the data provided by the authors as Suplementary Table 1 and load it into a dataframe. The dataset contains for example the MSI/MSS classifications.
```{r, warning=FALSE}
# suppress warning message caused by na values
dataset <- read_excel("C:/Users/Martin/Documents/Master/Forschungspraktikum/Chan/data/supplementary_table.xlsx")
# transform "NA" strings to na values
dataset[dataset=="NA"] <- NA 
```

We visualize the ratio of cell lines classified as MSS cell and cell lines classified as MSI cells for both Achilles and DRIVE datasets together and separately. We see that the number of MSS cell lines is much higher than the number of MSI cell lines.
```{r}
# count rows of MSI and MSS classified cell lines
counts_msi <- sum(dataset$CCLE_MSI=="MSI" & (!is.na(dataset$DRIVE_WRN_D2) | !is.na(dataset$CRISPR_WRN_CERES)), na.rm = TRUE)
counts_mss <- sum(dataset$CCLE_MSI=="MSS" & (!is.na(dataset$DRIVE_WRN_D2) | !is.na(dataset$CRISPR_WRN_CERES)), na.rm = TRUE)

# create dataframe with counts
counts.df <- data.frame(counts=c(counts_msi, counts_mss), classification=c("MSI", "MSS"))

# create plot for both datasets together
bar_chart <- ggplot(data=counts.df, aes(x=classification, y=counts)) +
  geom_bar(stat = "identity", width = 0.5) +
  ggtitle("Both Datasets in total")
bar_chart <- bar_chart + labs(x="",y="number of cell lines") + ylim(0,600)

# count rows of MSI and MSS classified cell lines in Achilles dataset
counts_msi_achilles <- sum(dataset$CCLE_MSI=="MSI" & !is.na(dataset$CRISPR_WRN_CERES), na.rm = TRUE)
counts_mss_achilles <- sum(dataset$CCLE_MSI=="MSS" & !is.na(dataset$CRISPR_WRN_CERES), na.rm = TRUE)
counts_achilles.df <- data.frame(counts=c(counts_msi_achilles, counts_mss_achilles), classification=c("MSI", "MSS"))

# create plot for Achilles dataset
bar_chart_achilles <- ggplot(data=counts_achilles.df, aes(x=classification, y=counts)) +
  geom_bar(stat = "identity", width = 0.5) +
  ggtitle("Achilles")
bar_chart_achilles <- bar_chart_achilles + labs(x="",y="number of cell lines") + ylim(0,600)

# count rows of MSI and MSS classified cell lines in DRIVE dataset
counts_msi_drive <- sum(dataset$CCLE_MSI=="MSI" & !is.na(dataset$DRIVE_WRN_D2), na.rm = TRUE)
counts_mss_drive <- sum(dataset$CCLE_MSI=="MSS" & !is.na(dataset$DRIVE_WRN_D2), na.rm = TRUE)
counts_drive.df <- data.frame(counts=c(counts_msi_drive, counts_mss_drive), classification=c("MSI", "MSS"))

# create plot for DRIVE dataset
bar_chart_drive <- ggplot(data=counts_drive.df, aes(x=classification, y=counts)) +
  geom_bar(stat = "identity", width = 0.5) +
  ggtitle("DRIVE")
bar_chart_drive <- bar_chart_drive + labs(x="",y="number of cell lines") + ylim(0,600)

# show plots
bar_chart + bar_chart_achilles + bar_chart_drive
```