---
title: "Data Exploration and Data Analysis"
subtitle: "Data Exploration and Implementation of the Workflow in 'WRN helicase is a synthetic lethal target in
microsatellite unstable cancers'"
author: "Martin Brand"
date: "`r format(Sys.time(), '%B %e, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    fig_caption: yes
    number_sections: no
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
    toc_collapsed: yes
    fig_width: 8 
    fig_height: 4 
  bookdown::pdf_document2:
    keep_tex: yes
    fig_caption: yes
    number_sections: no
    toc: yes
    toc_depth: 3
graphics: yes
header-includes:
- \makeatletter\renewcommand*{\fps@figure}{h}\makeatother
- \usepackage{placeins}
geometry: margin=1in
fontsize: 18pt
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE, results="hide", echo=TRUE}
library(ggplot2)
library(readxl)
library(patchwork)
library(tidyverse)
library(caret)
library(reshape2)
library(ggpubr)
library(ggrepel)
library(ggVennDiagram)
library(limma)
library(matrixTests)
```

## Preprocessing of the Data
I import the data provided by the authors as "Supplementary Table 1". The data set contains, among other things, the MSI/MSS classifications and the dependency scores for WRN. In addition, I import the data set "DepMap_18Q4_data.rds", which contains all dependency scores from the project Achilles and project DRIVE. The weblinks can all be found in @chan.
```{r, warning=FALSE}
# suppress warning message caused by NA values
data_dir <- "C:/Users/Martin/Documents/Master/Forschungspraktikum/genetic_dependencies_msi/data/"
# read file with MSS/MSI classifications
data_suppl <- read_excel(paste0(data_dir, "supplementary_table.xlsx"))
# transform "NA" strings to NA values
data_suppl[data_suppl=="NA"] <- NA 
# transform numbers saved as strings to numbers
data_suppl <- transform(data_suppl, DRIVE_WRN_D2=as.numeric(DRIVE_WRN_D2), CRISPR_WRN_CERES = as.numeric(CRISPR_WRN_CERES), avg_WRN_dep=as.numeric(avg_WRN_dep))
# delete rows which appear neither in Achilles nor in DRIVE dataset
data_suppl <- data_suppl[!is.na(data_suppl$DRIVE_WRN_D2) | !is.na(data_suppl$CRISPR_WRN_CERES), ]
# delete rows which are classified as intermediate
data_suppl <- data_suppl[data_suppl$CCLE_MSI!="indeterminate", ]
# delete rows which are NA in classification
data_suppl <- data_suppl[!is.na(data_suppl$CCLE_MSI), ]
# read file with dependency scores
data_18Q4 <- read_rds(file.path(data_dir, 'DepMap_18Q4_data.rds'))
dep_scores_18Q4_drive <- data_18Q4[["DRIVE"]]
dep_scores_18Q4_achilles <- data_18Q4[["CRISPR"]]
```

## Visualization of Classifications
I visualize the ratio of cell lines classified as MSS cell and cell lines classified as MSI cells for both Achilles and DRIVE data sets together and separately. We see that the number of MSS cell lines is many times higher than the number of MSI cell lines.
```{r}
# count rows of MSI and MSS classified cell lines
counts_class <- data.frame(table(data_suppl$CCLE_MSI))

# count rows of MSI and MSS classified cell lines for each dataset separately
counts_class_drive <- data.frame(table(data_suppl[!is.na(data_suppl$DRIVE_WRN_D2),]$CCLE_MSI))
counts_class_achilles <- data.frame(table(data_suppl[!is.na(data_suppl$CRISPR_WRN_CERES),]$CCLE_MSI))

# create plots
# define limits on y axis
limits <- c(0,600)

bar_chart_total <- ggplot(data=counts_class, aes(x=Var1, y=Freq, fill=Var1)) +
  theme(legend.position = "none") +
  geom_bar(stat = "identity", width = 0.5) +
  ggtitle("Both Datasets in total") +
  labs(x="",y="number of cell lines") +
  coord_cartesian(ylim = limits)

bar_chart_drive <- ggplot(data=counts_class_drive, aes(x=Var1, y=Freq, fill=Var1)) +
  theme(legend.position = "none") +
  geom_bar(stat = "identity", width = 0.5) +
  ggtitle("DRIVE") +
  labs(x="",y="number of cell lines") +
  coord_cartesian(ylim = limits)

bar_chart_achilles <- ggplot(data=counts_class_achilles, aes(x=Var1, y=Freq, fill=Var1)) +
  theme(legend.position = "none") +
  geom_bar(stat = "identity", width = 0.5) +
  ggtitle("Achilles") +
  labs(x="",y="number of cell lines") +
  coord_cartesian(ylim = limits)

bar_chart_total + bar_chart_drive + bar_chart_achilles
```

In the Figure above, we see that the datasets overlap. To determine the size of the overlap, I plot the data sets in a Venn diagram.
```{r}
DRIVE <- data_suppl[!is.na(data_suppl$DRIVE_WRN_D2),]$CCLE_ID
Achilles <- data_suppl[!is.na(data_suppl$CRISPR_WRN_CERES),]$CCLE_ID
all <- list(DRIVE=DRIVE, Achilles=Achilles)
ggVennDiagram(all, label_alpha = 0) +
  ggplot2::scale_fill_gradient(low="lightblue",high = "yellow") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  ggtitle('Cell Lines in Achilles and DRIVE Dataset')
```

## Correlation of Achilles and DRIVE Dataset
For quality control, I plot the dependency scores of WRN in the two data sets against each other. There is a significant correlation, but not as high as I would have expected. For this reason, I decided to analyse the two data sets separately instead of using the mean values. 
```{r}
# extract dependency scores
dep_scores <- data_suppl %>% select(5,6)
# remove rows with NA
dep_scores <- na.omit(dep_scores)
# create scatter plot
ggplot(dep_scores, aes(x=DRIVE_WRN_D2, y=CRISPR_WRN_CERES)) +
  geom_point() + stat_cor() +
  ggtitle('Complete Correlation Plot')
```

## Dependency Score Distribution in MSI and MSS Cell Lines
I plot the distributions of the dependency scores for WRN in MSS and MSI cells separately as histograms. The dependency scores in MSI cell lines show a significantly greater variance.
```{r}
# create MSI/MSS specific subsets with dependency scores
dep_scores <- data_suppl %>% select(1,4,5,6)
colnames(dep_scores) <- c("CCLE_ID", "CCLE_MSI", "DRIVE", "Achilles")
dep_scores <- melt(data = dep_scores, id.vars = c("CCLE_ID","CCLE_MSI"))
colnames(dep_scores) <- c("CCLE_ID", "CCLE_MSI", "dataset", "dep_score")
dep_scores <- dep_scores[!is.na(dep_scores$dep_score),]
dep_scores_mss <- data.frame(dep_scores[dep_scores$CCLE_MSI=="MSS",])
dep_scores_msi <- data.frame(dep_scores[dep_scores$CCLE_MSI=="MSI",])

# create plots for MSS and MSI cells separately
limits <- c(-3,1) # add limits in x axis

plot_mss <- ggplot(data = dep_scores_mss, aes(x=dep_score, fill=dataset)) +
  geom_histogram(binwidth = 0.2, position = "dodge") +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("Dependency Scores in MSS Cells") +
  theme(legend.position="bottom")

plot_msi <- ggplot(data = dep_scores_msi, aes(x=dep_score, fill=dataset)) +
  geom_histogram(binwidth = 0.2, position = "dodge") +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("Dependency Scores in MSI Cells") +
  theme(legend.position="bottom")

plot_mss + plot_msi
```

For better understanding, I repeat this step with density plots.
```{r}
plot_mss <- ggplot(data = dep_scores_mss, aes(x=dep_score, color=dataset)) +
  geom_density(linewidth=1, adjust = 0.3) +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("Dependency Scores in MSS Cells") +
  theme(legend.position="bottom") +
  geom_vline(xintercept = 0, linetype=2)

plot_msi <- ggplot(data = dep_scores_msi, aes(x=dep_score, color=dataset)) +
  geom_density(linewidth=1, adjust = 0.3) +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("Dependency Scores in MSI Cells") +
  theme(legend.position="bottom") +
  geom_vline(xintercept = 0, linetype=2)

plot_mss + plot_msi
```

For better understanding, I repeat this step with box plots and additionally calculate the p-values that are significantly low for both datasets.
```{r, echo=FALSE, message=FALSE, results="hide"}
# extract DRIVE and Achilles data separately with classifications
dep_scores_drive_with_class <- na.omit(data_suppl %>% select(4,5))
dep_scores_achilles_with_class <- na.omit(data_suppl %>% select(4,6))

box_plot_drive <- ggplot(dep_scores_drive_with_class, aes(x = CCLE_MSI, y = DRIVE_WRN_D2, fill = CCLE_MSI)) +
  geom_violin() +
  ggtitle("Boxplot for DRIVE") +
  theme(legend.position = "none") +
  geom_boxplot(outlier.shape = NA, show.legend = F,width = 0.2, position = position_dodge(width = 0.9)) +
  stat_compare_means(method = "wilcox.test")

box_plot_achilles <- ggplot(dep_scores_achilles_with_class, aes(x = CCLE_MSI, y = CRISPR_WRN_CERES, fill = CCLE_MSI)) +
  theme(legend.position = "none") +
  geom_violin() +
  ggtitle("Boxplot for Achilles") +
  geom_boxplot(outlier.shape = NA, show.legend = F,width = 0.2, position = position_dodge(width = 0.9)) +
  stat_compare_means(method = "wilcox.test")

box_plot_drive + box_plot_achilles
```

## Comparing WRN and ABHD1 Distribution
For comparison, I plot the distributions of dependency scores of a randomly selected KO gene and the WRN gene in the Achilles dataset as histograms. The dependency scores of WRN are generally more broadly distributed. When looking at MSI cell lines, the difference becomes much clearer.
```{r}
# extract dependency scores for ABHD1
abhd1_dep_scores_achilles <- data.frame(dep_scores_18Q4_achilles[,"ABHD1"])

# format dataset and add column for CCLE_ID
abhd1_dep_scores_achilles$CCLE_ID <- rownames(abhd1_dep_scores_achilles)
colnames(abhd1_dep_scores_achilles) <- c("dep_score_abhd1","CCLE_ID")
rownames(abhd1_dep_scores_achilles) <- c()

# extract subset containing CCLE_IDs, dependency scores and MSI/MSS classifications
dep_scores_achilles <- data_suppl %>% select(1,4,6)

# merge with dependency scores of ABHD1
dep_scores_achilles <- merge(abhd1_dep_scores_achilles, dep_scores_achilles, by="CCLE_ID")
dep_scores_achilles <- dep_scores_achilles[,c(1,3,2,4)]
colnames(dep_scores_achilles) <- c("CCLE_ID","CCLE_MSI","ABHD1", "WRN")

# melt dataset for better plotting
dep_scores_achilles <- melt(data = dep_scores_achilles,id.vars = c("CCLE_ID","CCLE_MSI"))
colnames(dep_scores_achilles) <- c("CCLE_ID","CCLE_MSI","KD_KO_gene", "dep_score")

# remove rows with NAs
dep_scores_achilles <- na.omit(dep_scores_achilles)

# create density plot for MSS cells
plot_mss <- ggplot(data=dep_scores_achilles[dep_scores_achilles$CCLE_MSI=="MSS",],aes(x=dep_score, fill=KD_KO_gene)) +
  geom_histogram(binwidth = 0.05, position = "dodge") +
  ggtitle("MSS in Achilles") +
  xlab("Dependency Score") +
  theme(legend.position="bottom")

# create density plot for MSI cells
plot_msi <- ggplot(data=dep_scores_achilles[dep_scores_achilles$CCLE_MSI=="MSI",],aes(x=dep_score, fill=KD_KO_gene)) +
  geom_histogram(binwidth = 0.05, position = "dodge") +
  ggtitle("MSI in Achilles") +
  xlab("Dependency Score") +
  theme(legend.position="bottom")

# plot results
plot_mss + plot_msi
```
I repeat the procedure using box plots and additionally calculate the p-values for comparison. The p-value for WRN is very significant, whereas the p-value for ABHD1 is not.
```{r}
dep_scores_achilles_wrn <- dep_scores_achilles[dep_scores_achilles$KD_KO_gene=="WRN",]
dep_scores_achilles_abhd1 <- dep_scores_achilles[dep_scores_achilles$KD_KO_gene=="ABHD1",]

box_plot_wrn <- ggplot(dep_scores_achilles_wrn, aes(x = CCLE_MSI, y = dep_score, fill = CCLE_MSI)) +
  geom_violin() +
  ggtitle("Boxplot for WRN") +
  theme(legend.position = "none") +
  geom_boxplot(outlier.shape = NA, show.legend = F,width = 0.2, position = position_dodge(width = 0.9)) +
  stat_compare_means(method = "wilcox.test")

box_plot_abhd1 <- ggplot(dep_scores_achilles_abhd1, aes(x = CCLE_MSI, y = dep_score, fill = CCLE_MSI)) +
  theme(legend.position = "none") +
  geom_violin() +
  ggtitle("Boxplot for ABHD1") +
  geom_boxplot(outlier.shape = NA, show.legend = F,width = 0.2, position = position_dodge(width = 0.9)) +
  stat_compare_means(method = "wilcox.test")

box_plot_wrn + box_plot_abhd1
```

# Calculating Mean Dependency Scores
To reproduce the workflow in @chan, I calculate the mean values of the dependency sores for MSI and MSS cell lines separately for every gene.
```{r}
# transform matrix to dataframe
dep_scores_18Q4_achilles_df <- data.frame(dep_scores_18Q4_achilles)

# add column with CCLE ID
dep_scores_18Q4_achilles_df <- cbind(CCLE_ID = rownames(dep_scores_18Q4_achilles_df), dep_scores_18Q4_achilles_df)
rownames(dep_scores_18Q4_achilles_df) <- NULL

# add CCLE_MSI classification to each cell line
classifications <- data_suppl %>% select(1,4)
dep_scores_18Q4_achilles_df <- merge(dep_scores_18Q4_achilles_df, classifications, by="CCLE_ID")

# split dataset in MSI and MSS
dep_scores_18Q4_achilles_msi <- dep_scores_18Q4_achilles_df[dep_scores_18Q4_achilles_df$CCLE_MSI=="MSI",c(-1,-ncol(dep_scores_18Q4_achilles_df))]
dep_scores_18Q4_achilles_mss <- dep_scores_18Q4_achilles_df[dep_scores_18Q4_achilles_df$CCLE_MSI=="MSS",c(-1,-ncol(dep_scores_18Q4_achilles_df))]

# calculate means
means_msi <- data.frame(colMeans(dep_scores_18Q4_achilles_msi))
means_mss <- data.frame(colMeans(dep_scores_18Q4_achilles_mss))

# merge datasets
means_msi$gene <- rownames(means_msi)
rownames(means_msi) <- NULL
means_mss$gene <- rownames(means_mss)
rownames(means_mss) <- NULL
mean_dep_scores <- merge(means_msi, means_mss, by="gene")
colnames(mean_dep_scores) <- c("gene","mean_msi","mean_mss")
head(mean_dep_scores)
```


## Calculating P-Values using Welch-Test
I calculate the p-values of the dependency scores for MSI and MSS cell lines for every gene using the Welch-Test. (Actually, the Welch-Test is not suitable for this data, but I do it for comparison.)
```{r}
res_welch <- col_t_welch(dep_scores_18Q4_achilles_msi, dep_scores_18Q4_achilles_mss)
p_values_welch <- data.frame(matrix(NA, nrow = nrow(res_welch), ncol = 2))
colnames(p_values_welch) <- c('gene','p_value')
p_values_welch$gene <- rownames(res_welch)
p_values_welch$p_value <- res_welch$pvalue
head(p_values_welch)
```

## Mean Differences against P-Values using Welch-Test
I plot the mean differences of the dependency scores in MSI and MSS cell lines against the corresponding p-values. The dependency scores of WRN show a high mean difference and a significant p-value.
```{r}
mean_differences <- mean_dep_scores
mean_differences$mean_msi <- NULL
mean_differences$mean_mss <- NULL
mean_differences$difference <- (mean_dep_scores$mean_msi - mean_dep_scores$mean_mss)

# transform p-values
p_adj_welch <- p_values_welch
p_adj_welch$p_value <- p.adjust(p_values_welch$p_value, method = "hochberg")
p_values_welch_log <- p_adj_welch
p_values_welch_log$p_value <- NULL
p_values_welch_log$log <- -log(p_adj_welch$p_value)

# create data structure for mean differences and p-values
data <- merge(mean_differences, p_values_welch_log, by="gene")

# define labeled genes
data$to_label <- ""
data$to_label[data$gene == "WRN"] <- "WRN"

ggplot(data=data, aes(x=difference, y=log, label=to_label)) +
  geom_point() +
  xlab('MSI–MSS mean difference') +
  ylab('-log(P)') +
  geom_label_repel()
```

## Calculating P-Values using Wilcoxon Test
Again, I calculate the p-values of the dependency scores for MSI and MSS cell lines for every gene, but using the Wilcoxon-Test.
```{r}
res_wilc <- col_wilcoxon_twosample(dep_scores_18Q4_achilles_msi, dep_scores_18Q4_achilles_mss)
p_values_wilc <- data.frame(matrix(NA, nrow = nrow(res_wilc), ncol = 2))
colnames(p_values_wilc) <- c('gene','p_value')
p_values_wilc$gene <- rownames(res_wilc)
p_values_wilc$p_value <- res_wilc$pvalue
head(p_values_wilc)
```

## Mean Differences against P-Values using Wilcoxon-Test
Again, I plot the mean differences of the dependency scores in MSI and MSS cell lines against the corresponding p-values. Using the appropriate test method, the p-value for WRN is even higher and the result is therefor more reliable.
```{r}
# transform p-values
p_adj_wilc <- p_values_wilc
p_adj_wilc$p_value <- p.adjust(p_values_wilc$p_value, method = "hochberg")
p_values_wilc_log <- p_adj_wilc
p_values_wilc_log$p_value <- NULL
p_values_wilc_log$log <- -log(p_adj_wilc$p_value)

# create data structure for mean differences and p-values
data <- merge(mean_differences, p_values_wilc_log, by="gene")

# define labeled genes
data$to_label <- ""
data$to_label[data$gene == "WRN"] <- "WRN"

ggplot(data=data, aes(x=difference, y=log, label=to_label)) +
  geom_point() +
  xlab('MSI–MSS mean difference') +
  ylab('-log(P)') +
  geom_label_repel()
```

## Volcano Plot using Bayes moderated t-stats
Now I calculate the p-values using Bayes moderated t-stats in limma [@limma] as in @chan and plot them against the corresponding log2FoldChanges. With this recommended test method, the result is even more reliable.
```{r}
vec <- dep_scores_18Q4_achilles_df$CCLE_MSI
mat <- dep_scores_18Q4_achilles_df
mat$CCLE_MSI <- NULL
rownames(mat) <- mat$CCLE_ID
mat$CCLE_ID <- NULL
mat <- as.matrix(mat)

design <- model.matrix(~vec)
fit <- lmFit(t(mat),design=design)
fit <- eBayes(fit = fit)
results <- topTable(fit = fit, n=Inf, coef = 2)

# define labeled genes
results$to_label <- ""
results$to_label[rownames(results) == "WRN"] <- "WRN"

ggplot(data=results, aes(x=logFC,y=-log(adj.P.Val), label=to_label)) +
  geom_point() +
  xlab('logFoldChange') +
  ylab('-log(P_adj)') +
  geom_label_repel()
```