---
title: "Data Exploration"
subtitle: "Data Exploration of 'WRN helicase is a synthetic lethal target in
microsatellite unstable cancers'"
author: "Martin Brand"
date: "`r format(Sys.time(), '%B %e, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    fig_caption: yes
    number_sections: no
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
    toc_collapsed: yes
    fig_width: 8 
    fig_height: 4 
  bookdown::pdf_document2:
    keep_tex: yes
    fig_caption: yes
    number_sections: no
    toc: yes
    toc_depth: 3
graphics: yes
header-includes:
- \makeatletter\renewcommand*{\fps@figure}{h}\makeatother
- \usepackage{placeins}
geometry: margin=1in
fontsize: 18pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE, results="hide", echo=TRUE}
library(ggplot2)
library(readxl)
library(patchwork)
library(tidyverse)
library(caret)
library(reshape2)
library(ggpubr)
```

## Preprocessing of the Data
We import the data provided by the authors as "Supplementary Table 1" and load it into a dataframe. The dataset contains for example the MSI/MSS classifications. Additionally, we import the "DepMap_18Q4_data.rds" dataset, which contains all dependency scores. 
```{r, warning=FALSE}
# suppress warning message caused by na values
data_dir <- "C:/Users/Martin/Documents/Master/Forschungspraktikum/Chan/data/"
# read file with MSS/MSI classifications
data_suppl <- read_excel(paste0(data_dir, "supplementary_table.xlsx"))
# transform "NA" strings to na values
data_suppl[data_suppl=="NA"] <- NA 
# transform numbers saved as strings to numbers
data_suppl <- transform(data_suppl, DRIVE_WRN_D2=as.numeric(DRIVE_WRN_D2), CRISPR_WRN_CERES = as.numeric(CRISPR_WRN_CERES), avg_WRN_dep=as.numeric(avg_WRN_dep))
# read file with dependency scores
data_18Q4 <- read_rds(file.path(data_dir, 'DepMap_18Q4_data.rds'))
dep_scores_18Q4_drive <- data_18Q4[["DRIVE"]]
dep_scores_18Q4_achilles <- data_18Q4[["CRISPR"]]
```

## Visualization of Classifications
We visualize the ratio of cell lines classified as MSS cell and cell lines classified as MSI cells for both Achilles and DRIVE datasets together and separately. We see that the number of MSS cell lines is much higher than the number of MSI cell lines.
```{r}
# count rows of MSI and MSS classified cell lines
counts_msi <- sum(data_suppl$CCLE_MSI=="MSI" & (!is.na(data_suppl$DRIVE_WRN_D2) | !is.na(data_suppl$CRISPR_WRN_CERES)), na.rm = TRUE)
counts_mss <- sum(data_suppl$CCLE_MSI=="MSS" & (!is.na(data_suppl$DRIVE_WRN_D2) | !is.na(data_suppl$CRISPR_WRN_CERES)), na.rm = TRUE)

# create dataframe with counts
counts.df <- data.frame(counts=c(counts_msi, counts_mss), classification=c("MSI", "MSS"))

# define limits on y axis
limits <- c(0,600)

# create plot for both datasets together
bar_chart <- ggplot(data=counts.df, aes(x=classification, y=counts)) +
  geom_bar(stat = "identity", width = 0.5) +
  ggtitle("Both Datasets in total")
bar_chart <- bar_chart + labs(x="",y="number of cell lines") + coord_cartesian(ylim = limits)

# count rows of MSI and MSS classified cell lines in Achilles dataset
counts_msi_achilles <- sum(data_suppl$CCLE_MSI=="MSI" & !is.na(data_suppl$CRISPR_WRN_CERES), na.rm = TRUE)
counts_mss_achilles <- sum(data_suppl$CCLE_MSI=="MSS" & !is.na(data_suppl$CRISPR_WRN_CERES), na.rm = TRUE)
counts_achilles.df <- data.frame(counts=c(counts_msi_achilles, counts_mss_achilles), classification=c("MSI", "MSS"))

# create plot for Achilles dataset
bar_chart_achilles <- ggplot(data=counts_achilles.df, aes(x=classification, y=counts)) +
  geom_bar(stat = "identity", width = 0.5) +
  ggtitle("Achilles")
bar_chart_achilles <- bar_chart_achilles + labs(x="",y="number of cell lines") + coord_cartesian(ylim = limits)

# count rows of MSI and MSS classified cell lines in DRIVE dataset
counts_msi_drive <- sum(data_suppl$CCLE_MSI=="MSI" & !is.na(data_suppl$DRIVE_WRN_D2), na.rm = TRUE)
counts_mss_drive <- sum(data_suppl$CCLE_MSI=="MSS" & !is.na(data_suppl$DRIVE_WRN_D2), na.rm = TRUE)
counts_drive.df <- data.frame(counts=c(counts_msi_drive, counts_mss_drive), classification=c("MSI", "MSS"))

# create plot for DRIVE dataset
bar_chart_drive <- ggplot(data=counts_drive.df, aes(x=classification, y=counts)) +
  geom_bar(stat = "identity", width = 0.5) +
  ggtitle("DRIVE")
bar_chart_drive <- bar_chart_drive + labs(x="",y="number of cell lines") + coord_cartesian(ylim = limits)

# show plots
bar_chart + bar_chart_achilles + bar_chart_drive
```

## Correlation of Achilles and DRIVE dataset
```{r}
# extract dependency scores
dep_scores <- data_suppl %>% select(5,6)
# remove rows with na
dep_scores <- na.omit(dep_scores)
# # normalise values
# process <- preProcess(dep_scores, method=c("range"))
# norm_scores <- predict(process, dep_scores)

# remove outliers
outliers_achilles <- boxplot(dep_scores$CRISPR_WRN_CERES, plot=FALSE)$out
cleaned_dep_scores<- dep_scores[-which(dep_scores$CRISPR_WRN_CERES %in% outliers_achilles),]
outliers_drive <- boxplot(cleaned_dep_scores$DRIVE_WRN_D2, plot=FALSE)$out
cleaned_dep_scores<- cleaned_dep_scores[-which(cleaned_dep_scores$DRIVE_WRN_D2 %in% outliers_drive),]
# create scatter plot with and without outliers
corr_plot <- ggplot(dep_scores, aes(x=DRIVE_WRN_D2, y=CRISPR_WRN_CERES)) +
  geom_point() + stat_cor()
cleaned_corr_plot <- ggplot(cleaned_dep_scores, aes(x=DRIVE_WRN_D2, y=CRISPR_WRN_CERES)) +
  geom_point() + stat_cor()
corr_plot + cleaned_corr_plot
```

## Histograms
We visualize the the difference between the dependency scores for MSS and MSI cells as histograms.
```{r}
# create MSI/MSS specific subsets with dependency scores
dep_scores_melted <- data_suppl %>% select(1,4,5,6)
dep_scores_melted <- dep_scores_melted[!is.na(dep_scores_melted$CCLE_MSI),]
colnames(dep_scores_melted) <- c("CCLE_ID", "CCLE_MSI", "DRIVE", "Achilles")
dep_scores_melted <- melt(data = dep_scores_melted, id.vars = c("CCLE_ID","CCLE_MSI"))
colnames(dep_scores_melted) <- c("CCLE_ID", "CCLE_MSI", "dataset", "dep_score")
dep_scores_melted <- dep_scores_melted[!is.na(dep_scores_melted$dep_score),]
dep_scores_mss <- data.frame(dep_scores_melted[dep_scores_melted$CCLE_MSI=="MSS",])
dep_scores_msi <- data.frame(dep_scores_melted[dep_scores_melted$CCLE_MSI=="MSI",])

# create plots for MSS and MSI cells separately
limits <- c(-3,1) # add limits in x axis

plot_mss <- ggplot(data = dep_scores_mss, aes(x=dep_score, fill=dataset)) +
  geom_histogram(binwidth = 0.2) +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("Dependency Scores in MSS Cells") +
  theme(legend.position="bottom")

plot_msi <- ggplot(data = dep_scores_msi, aes(x=dep_score, fill=dataset)) +
  geom_histogram(binwidth = 0.2) +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("Dependency Scores in MSI Cells") +
  theme(legend.position="bottom")

plot_mss + plot_msi

# the offset is different in DRIVE and Achilles... why?
```

## Density Plots
We visualize the the difference between the dependency scores for MSS and MSI cells as density plots.
```{r}
plot_mss <- ggplot(data = dep_scores_mss, aes(x=dep_score, color=dataset)) +
  geom_density(size=1) +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("Dependency Scores in MSS Cells") +
  theme(legend.position="bottom") +
  geom_vline(xintercept = 0, linetype=2)

plot_msi <- ggplot(data = dep_scores_msi, aes(x=dep_score, color=dataset)) +
  geom_density(size=1) +
  coord_cartesian(xlim = limits) +
  xlab("Dependency Score") +
  ggtitle("Dependency Scores in MSI Cells") +
  theme(legend.position="bottom") +
  geom_vline(xintercept = 0, linetype=2)

plot_mss + plot_msi
```

## Box Plots
We visualize the the difference between the dependency scores for MSS and MSI cells as box plots.
```{r}
# split Achilles and DRIVE datasets
dep_scores_mss_drive <- dep_scores_mss[dep_scores_mss$dataset=="DRIVE",]
dep_scores_mss_achilles <- dep_scores_mss[dep_scores_mss$dataset=="Achilles",]
dep_scores_msi_drive <- dep_scores_msi[dep_scores_msi$dataset=="DRIVE",]
dep_scores_msi_achilles <- dep_scores_msi[dep_scores_msi$dataset=="Achilles",]

df <- rbind(dep_scores_mss, dep_scores_msi)

box_plot <- ggplot(df, aes(x = dataset, y= dep_score, fill = CCLE_MSI)) +
  geom_violin() +
  geom_boxplot(outlier.shape = NA, show.legend = F,width = 0.2, position = position_dodge(width = 0.9))

box_plot
```

## Distribution of randomly picked KD/KO genes
For comparison we plot the distribution of randomly picked KO/KD genes.
```{r}
# extract dependency scores for ABHD1
abhd1_dep_scores_achilles <- data.frame(dep_scores_18Q4_achilles[,"ABHD1"])

# format dataset and add column for CCLE_ID
abhd1_dep_scores_achilles$CCLE_ID <- rownames(abhd1_dep_scores_achilles)
colnames(abhd1_dep_scores_achilles) <- c("dep_score_abhd1","CCLE_ID")
rownames(abhd1_dep_scores_achilles) <- c()

# extract subset containing CCLE_IDs, dependency scores and MSI/MSS classifications
dep_scores_achilles <- data_suppl %>% select(1,4,6)

# merge with dependency scores of ABHD1
dep_scores_achilles <- merge(abhd1_dep_scores_achilles, dep_scores_achilles, by="CCLE_ID")
dep_scores_achilles <- dep_scores_achilles[,c(1,3,2,4)]
colnames(dep_scores_achilles) <- c("CCLE_ID","CCLE_MSI","ABHD1", "WRN")

# melt dataset for better plotting
dep_scores_achilles <- melt(data = dep_scores_achilles,id.vars = c("CCLE_ID","CCLE_MSI"))
colnames(dep_scores_achilles) <- c("CCLE_ID","CCLE_MSI","KD_KO_gene", "dep_score")

# remove rows with NAs
dep_scores_achilles <- na.omit(dep_scores_achilles)

# create density plot for MSS cells
plot_mss <- ggplot(data=dep_scores_achilles[dep_scores_achilles$CCLE_MSI=="MSS",],aes(x=dep_score, fill=KD_KO_gene)) +
  geom_histogram(binwidth = 0.05) +
  ggtitle("MSS in Achilles") +
  xlab("Dependency Score") +
  theme(legend.position="bottom")

# create density plot for MSI cells
plot_msi <- ggplot(data=dep_scores_achilles[dep_scores_achilles$CCLE_MSI=="MSI",],aes(x=dep_score, fill=KD_KO_gene)) +
  geom_histogram(binwidth = 0.05) +
  ggtitle("MSI in Achilles") +
  xlab("Dependency Score") +
  theme(legend.position="bottom")

# plot results
plot_mss + plot_msi
```